---
title: "c7091_Problem_Set"
format: html
editor: visual
---

## C7091 Professional Skills for Data Science

Miss S-J Childs\
Student ID: 24370500\
Problem Set\
2024-10-23\

```{r setup, include = FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

install.packages("emmeans")
install.packages("pwr")
install.packages("tidyverse")
library(emmeans)
library(pwr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)

# Ensure the datasets are loaded
data("mpg")
data("iris")
data("OrchardSprays")
```

### Question One

"Edit the following code so that the plot accurately reflects the axis labels and the subset functions returns an object with only the contents intended (as shown in the preceding comment to the function). Using indexing and regular expression tools in base R, extend the code to calculate the proportions of manual and automatic shift cars in the study"

```{r}
# Corrected code
plot(mpg$displ, mpg$hwy, col = as.factor(mpg$year), 
     main = "Displacement by Highway Miles per Gallon", # added title
     xlab = "Displacement (L)", 
     ylab = "Highway (MPG)", pch = 19) # added measurement units
legend("topright", legend = unique(mpg$year), 
       col = unique(as.factor(mpg$year)), pch = 19)
```

```{r}
# Corrected code
subset(mpg, cyl == 8) 
# 8 is a numeric, not a character. == for equality, = means assignment
```

```{r}
# Created a table of transmission types
trans_table <- table(mpg$trans) # shows the frequency of different types

# Calculated the proportions
#  help("proportions") # entries of x, divided by the appropriate marginal sums

trans_proportions <- proportions(trans_table) # but is this helpful?
View(trans_proportions)
```

```{r }
# Visualization of transmission types
barplot(trans_table, main ="Transmission Types in the mpg Dataset", 
        xlab ="Transmission Type",
        ylab ="Frequency", col="blue") # now easier to comprehend proportions
```

### Question Two

"Write pseudocode steps for calculating the volume of a cylinder (hint, if you do not know it by heart, you may need to research the equation for the volume of a cylinder!). For a cylinder of height = 3.2 cm and end radius of 5.5 cm, report the volume in cm to 2 decimal points of accuracy. Use at least 3 decimal points of accuracy for pi (hint, the quantity named pi is a standard variable in R!)"

```{r}
# Step One: Look up the formula for cylinder volume
# Volume = π * radius^2 * height

# Step Two: State the height and radius for cylinder
height <- 3.2
radius <- 5.5

# Step Three: State value of pi to 3 decimal places
print(pi) # checking
pie <- round(pi, 3)
print(pie) # checking

# Step Four: Use formula to calculate 
cy_volume <- pie * radius^2 * height

# Step Five: Round the volume to 2 decimal points
cy_volume_rounded <- round(cy_volume, 2)

# Step Six: Show the answer
print(cy_volume_rounded)
```

### Question Three

"In your own words, what value is required for the d argument in the pwr.t.test() function in the {pwr} package? Show the code involved including any appropriate comment code required to answer this question. (hint: you will probably need to install the package, load it, and use help() on the function name)"

```{r}
# A statistical method known as power analysis. 
# My understanding of power analysis as having four core elements:
# Statistical power == detecting an effect if there's one
# Significance aka 'alpha' == the level needed (0.05) to reject the null hypothesis
# Sample size == the number of observations
# Effect size == the difference between the mean divided by the combined
# standard deviation, aka 'Cohen's d' and the 'd' argument in pwr.t.test function.
# The 'd' argument is the mean difference / standard deviation difference

# Example code
# mean_diff <- mean("hypothetical data B" - "hypothetical data A")
# sd_diff <- sd("hypothetical data B" - "hypothetical data A")
# d <- mean_diff / sd_diff
```

### Question Four

"Using the code chunk below, please answer the following questions:

What is the role of the set.seed() function in the context of this code? Why does calculating the mean of the copied original matrix return a numeric value while the mean of the same matrix when missing values are filled from the list returns NA? Please fix the code such that mean_of_original_copy and mean_matrix are equal"

```{r}
# help("set.seed")
# It's a random number generator and helps with reproducibility.
# The question code chunk has random sampling, but you don't 
# want the results to change each time you run the code, so 
# you set a seed.
```

In R you need to be careful when making copies. Making a copy of the matrix, but then altering the original will NOT change the copied version. This is because when copying a matrix, R creates a new independent copy of the data.

```{r}
# Corrected code

set.seed(123)
matrix_with_na <- matrix(sample(c(NA, 1:9), 9, replace = TRUE), nrow = 3)
copy_of_original<-matrix_with_na

matrix_with_na[1,2] <- NA
matrix_with_na[2,3] <- NA
matrix_with_na[3,1] <- NA

print(copy_of_original)
print(matrix_with_na)

list_with_values <- as.list(as.numeric(matrix(1:9, nrow = 3)))
# originally as.character, must be numeric or logical

print(list_with_values)

for (i in 1:3) {
  for (j in 1:3) {
    if (is.na(matrix_with_na[i,j])) {
      matrix_with_na[i,j] <- list_with_values[[(i-1)*3 + j]] 
    }
  }
}
print(matrix_with_na)

# Now calculating the mean of the matrices
mean_matrix <- mean(matrix_with_na)
mean_of_original_copy <- mean(copy_of_original)
print(mean_of_original_copy)
print(mean_matrix)

# Both matrices now match at 4.333333
```

### Question Five

"Explain why this function is not working and return the working equivalent. Interpret what the estimate means and rank the species"

```{r}
data("iris")

# Original code leads to "error: arguments imply differing number of rows: 0, 3"

names(iris) # Typos in original, checking names

# Corrected code
anova_result <- aov(Sepal.Length ~ Species, data = iris)
run_anova_posthoc <- function() {
  posthoc_result <- emmeans(anova_result, pairwise ~ Species)
  comparisons <- as.data.frame(summary(posthoc_result$contrast)) # for accessibility
  print(comparisons)
  comparisons_df <- data.frame(Species = comparisons$contrast, # NOT $species
                               estimate = comparisons$estimate,
                               p.value = comparisons$p.value)
  return(comparisons_df)
}

run_anova_posthoc()
emmeans(anova_result, ~ Species) # check the emmean column to rank


```
